---
title: "An introduction to GAM(M)s"
author: "Stefano Coretta"
date: "12/07/2018"
output:
  beamer_presentation:
    includes:
      in_header: header.tex
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(mgcv)
library(itsadug)
library(tidymv)
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")
scale_fill_discrete <- function(...) scale_fill_manual(values = cbPalette, ...)
scale_colour_discrete <- function(...) scale_colour_manual(values = cbPalette, ...)
scale_fill_viridis_d <- function(...) scale_fill_manual(values = cbPalette, ...)
scale_colour_viridis_d <- function(...) scale_colour_manual(values = cbPalette, ...)
```

```{r data, message=FALSE}
set.seed(8788)
x <- seq(0,100,1)
y <- ((runif(1, 10, 20) * x) / (runif(1, 0, 10) + x)) + rnorm(101, 0, 1)
x_1 <- seq(0, 100, 1)
y_1 <- ((runif(1, 10, 20) * x_1) / (runif(1, 0, 100) + x_1)) + rnorm(101, 0, 1)

sim_nl <- tibble(
  x = c(x, x_1),
  y = c(y, y_1),
  group = rep(c("a", "b"), each = 101)
)

sim_nl_a <- filter(sim_nl, group == "a")

vowels <- read_csv("./data/vowels.csv") %>%
  mutate(
    vowel = as.ordered(vowel),
    voicing = ordered(voicing, levels = c("voiceless", "voiced")),
    place = as.ordered(place),
    vow_voi = interaction(vowel, voicing),
    vow_voi = as.ordered(vow_voi),
    speaker = as.factor(speaker)
  )

contrasts(vowels$vowel) <- "contr.treatment"
contrasts(vowels$voicing) <- "contr.treatment"
contrasts(vowels$place) <- "contr.treatment"
contrasts(vowels$vow_voi) <- "contr.treatment"
```

## Linear models

\centering \Huge

Time travel...

## Linear models

\centering \Huge

$y = 3 + 2x$

\large

where $x = (2, 4, 5, 8, 10, 23, 36)$

## Linear models

```{r homework}
line <- tibble(
  x = c(2, 4, 5, 8, 10, 23, 36),
  y = 3 + 2 * x
)
ggplot(line, aes(x, y)) +
  geom_point() +
  labs(title = bquote(italic(y) == 3 + 2 * italic(x)))
```

## Linear models

```{r line}
ggplot(line, aes(x, y)) +
  geom_point(colour = "gray") +
  geom_line(colour = "red") +
  labs(title = bquote(italic(y) == 3 + 2 * italic(x)))
```

## Linear models

* In science, we have $x$ and $y$...

* for example, vowel duration and VOT, speech rate and pitch, etc...

## Linear models

```{r sampled}
set.seed(4321)
x <- 1:10
y <- (1 + 1.5 * x) + rnorm(10, 0, 1)

line <- tibble(
  x = x,
  y = y
)
ggplot(line, aes(x, y)) +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == "?" + "?" * italic(x)))
```

## Linear models

* The formula: $y = \beta_0 + \beta_1x$
    * $\beta_0$ is the **intercept**
    * $\beta_1$ is the **slope**

* We know $x$ and $y$
    * we need to estimate $\beta_0$, $\beta_1$ = $\hat{\beta_0}, \hat{\beta_1}$

* We can add more predictors
    * $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_nx_n$

* `lm(y ~ x, data)` ('$y$ as a function of $x$')

## Linear models

```{r lm-plot}
ggplot(line, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote({italic(y) == beta[0] + beta[1] * italic(x)} == 1 + 1.5 * italic(x)))
```

## Linear models

```{r error}
m <- lm(y ~ x)
yhat <- m$fitted.values
diff <- y - yhat  
ggplot(line, aes(x, y)) +
  geom_segment(aes(x = x, xend = x, y = y, yend = yhat), colour = "red") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + epsilon))
```

## LM with non-linear data

```{r plot-data}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point() +
  labs(title = "Some non-linear data")
```

## LM with non-linear data

```{r plot-lm}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Some non-linear data")
```

## LM with non-linear data

```{r error-2}
m <- lm(y ~ x, data = sim_nl_a)
yhat <- m$fitted.values
diff <- sim_nl_a$y - yhat  
ggplot(sim_nl_a, aes(x, y)) +
  geom_segment(aes(x = x, xend = x, y = y, yend = yhat), colour = "red") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Some non-linear data")
```

## LM with non-linear data

How to account for non-linearity in a linear model?

* Use **higher-degree polynomials**
    * quadratic: $y = \beta_0 + \beta_1x + \beta_2x^2$
    * cubic: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3$
    * $n$th: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3 + ... + \beta_nx^n$

## LM with non-linear data

```{r 2-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2))
```

## LM with non-linear data

```{r 3-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3))
```

## LM with non-linear data

```{r 10-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10))
```

## LM with non-linear data

```{r 10-poly-2, warning=FALSE}
ggplot(line, aes(x, y)) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10)) +
  ylim(1, 18)
```

## LM with non-linear data

```{r 10-poly-21, warning=FALSE}
ggplot(line, aes(x, y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10)) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10)) +
  ylim(1, 18)
```

## Generalised additive models

* **G**enrealised **A**dditive **M**odel**s**

* $y = f(x) + \epsilon$
    * $f(x)$ = 'some function of $x$' (or *smooth function*)

## Smooth terms

* LMs have **parametric terms**
    * $\beta_nx_n$ (`x` in `R`)
    * linear effects

* GAMs add (non-parametric) **smooth terms** (or simply smooths, also smoothers)
    * $f(x)$, `s(x)` in `R`
    * non-linear effects

* `gam(y ~ s(x), data)`, '$y$ as *some* function of $x$'

## Smoothing splines, basis, basis functions

* Smooths in GAMs are **smoothing splines**
    * splines are defined piecewise with a set of polynomials

* The set of polynomials is called a **basis**
    * the basis is composed of **basis functions** (the polynomials)

* A spline is the sum of the products of each basis function and its coefficient

## Basis functions

```{r basis}
par(mfrow = c(1, 2))

simple <- gam(y ~ s(x, bs = "cr", k = 10), data = sim_nl_a)

X <- predict(
  simple,
  newdata = data.frame(x = sim_nl_a$x),
  type = "lpmatrix"
)[,2:10]

plot(sim_nl_a$x, X[, 1], type = "n", ylim = c(-2, 3), xlab = "x", ylab = "y")
for (i in 1:ncol(X)) {
  lines(sim_nl_a$x, X[, i], col = i, lw = 1)
}

X2 <- X %*% diag(coef(simple)[2:10])
plot(
  sim_nl_a$x,
  predict(simple, newdata = data.frame(x = sim_nl_a$x)) -
    coef(simple)[1],
  type = "n",
  ylim = c(-8, 3),
  xlab = "x", ylab = "y"
)
for (i in 1:ncol(X2)) {
  lines(sim_nl_a$x, X2[, i], col = i, lw = 1)
}
lines(
  sim_nl_a$x,
  predict(simple, newdata = data.frame(x = sim_nl_a$x)) -
    coef(simple)[1],
  lw = 3
)
```

## Smoothing parameter

* 'Wiggliness' is related to number of basis functions
    * more basis functions, more wiggliness (less smoothing)

* The **smoothing parameter** penalises wiggliness
    * high values = less wiggliness (more smoothing)
    * estimated from the data

## Smoothing splines

* There are **several kinds** of splines
    * each with their own basis functions

* Most common
    * *thin plate regression splines*
    * *cubic regression splines*

* For more info, run `?smooth.terms`

## A simple GAM

\tiny

```{r simple-10, echo = TRUE}
simple <- gam(y ~ s(x, bs = "cr", k = 10), data = sim_nl_a)
summary(simple)
```

## A simple GAM

```{r simple-10-plot}
plot(simple)
```

## A simple GAM

```{r simple-5}
simple <- gam(y ~ s(x, bs = "cr", k = 5), data = sim_nl_a)
plot(simple)
```

## Comparing levels

```{r nl-plot}
ggplot(sim_nl, aes(x, y, colour = group)) +
  geom_point()
```

## Comparing levels

```{r nl-smooth, message=FALSE}
ggplot(sim_nl, aes(x, y, colour = group)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

## Comparing levels

* `by`-variables with ordered factors

```{r by-variables, echo = TRUE, eval=FALSE}
compare <- gam(
  y ~
    group +
    s(x, bs = "cr", k = 5) +
    s(x, bs = "cr", k = 5, by = group),
  data = sim_nl
)
```

## Comparing levels

* To use `by`-variables with ordered factors
    * change factor to **ordered factor**
    * change factor contrast to **treatment contrast** (`contr.treatment`)
        * the default in ordered factors is `contr.poly`, this won't work
    * include factor as **parametric term**
    * include a **reference smooth** and a **difference smooth** with the `by`-variable

## Comparing levels

```{r ordered, echo=TRUE}
sim_nl <- sim_nl %>%
  mutate(group = ordered(group, levels = c("a", "b")))
contrasts(sim_nl$group) <- "contr.treatment"
```

## Comparing levels

```{r compare-gam, echo = TRUE}
library(mgcv)

compare <- gam(
  y ~
    group +
    s(x, bs = "cr", k = 5) +
    s(x, bs = "cr", k = 5, by = group),
  data = sim_nl
)
```

## Comparing levels

\tiny

```{r compare-summary}
summary(compare)
```

## Comparing levels

```{r plot-smooth-fn, eval=FALSE, echo=TRUE}
library(tidymv)

plot_smooths(compare, x, group)
```

## Comparing levels

```{r plot-smooths}
plot_smooths(compare, x, group)
```

## Significance testing

* Several ways for testing significance of smooths

* We will used a combined method
    * model comparison with `itsadug::compareML()` of a full and a null model
    * visualisation of the difference smooth with `tidymv::plot_difference()`
        * (you can also use `itsadug::plot_diff()`)

## Significance testing

\footnotesize

```{r ml, echo=TRUE, eval=FALSE}
compare_1 <- gam(
  y ~
    group +
    s(x, bs = "cr", k = 5) +
    s(x, bs = "cr", k = 5, by = group),
  data = sim_nl,
  method = "ML"
)

compare_0 <- gam(
  y ~
    s(x, bs = "cr", k = 5),
  data = sim_nl,
  method = "ML"
)
```

## Significance testing

```{r compare-1-0}
compare_1 <- gam(
  y ~
    group +
    s(x, bs = "cr", k = 5) +
    s(x, bs = "cr", k = 5, by = group),
  data = sim_nl,
  method = "ML"
)

compare_0 <- gam(
  y ~
    s(x, bs = "cr", k = 5),
  data = sim_nl,
  method = "ML"
)
```

\footnotesize

```{r compare-ml, echo=TRUE}
compareML(compare_0, compare_1)
```

## Significance testing

* Let's plot the difference smooth with `tidymv::plot_difference()`

```{r difference, echo=TRUE, eval=FALSE}
plot_difference(compare, x, list(group = c("b", "a")))
```

## Significance testing

```{r plot-difference}
plot_difference(compare, x, list(group = c("b", "a")))
```

## Hands on

\Huge
\centering

Tutorial 1

## Autocorrelation

* more data

* use `bam()` (big GAM)

## Autocorrelation

```{r bam, echo=TRUE}
big_gam <- bam(
  f1 ~
    voicing +
    vowel +
    s(time_point, k = 6) +
    s(time_point, k = 6, by = voicing) +
    s(time_point, k = 6, by = vowel),
  data = vowels
)
```


## Autocorrelation

\tiny

```{r bam-summary}
summary(big_gam)
```

## Autocorrelation

```{r bam-plot}
plot_smooths(big_gam, time_point, voicing, vowel)
```

## Autocorrelation

```{r bam-acf}
acf_plot(resid(big_gam), split_by = list(vowels$index))
```

## Autocorrelation

```{r bam-start-event}
vowels <- vowels %>%
  arrange(speaker, index, time_point) %>%
  create_event_start("index")
```

## Autocorrelation

```{r ar-gam}
rho <- start_value_rho(big_gam)

ar_gam <- bam(
  f1 ~
    voicing +
    vowel +
    s(time_point, k = 6) +
    s(time_point, k = 6, by = voicing) +
    s(time_point, k = 6, by = vowel),
  data = vowels,
  rho = rho,
  AR.start = vowels$start.event
)
```

## Autocorrelation

```{r ar-gam-plot}
plot_smooths(ar_gam, time_point, voicing, vowel)
```

## Autocorrelation

```{r ar-gam-acf}
acf_resid(ar_gam, split_pred = "AR.start")
```

## Random effects

<!-- say random needs to be a factor -->

```{r random-gam}
random_gam <- bam(
  f1 ~
    voicing +
    vowel +
    s(time_point, k = 6) +
    s(time_point, k = 6, by = voicing) +
    s(time_point, k = 6, by = vowel) +
    s(time_point, speaker, bs = "fs", m = 1),
  data = vowels
)
summary(random_gam)
```

## Random effects

```{r random-gam-plot}
plot_smooths(random_gam, time_point, voicing, vowel)
```

## Interactions

```{r interactions}
ggplot(vowels, aes(time_point, f1, colour = voicing)) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 5)) +
  facet_grid(place ~ vowel)
```

## Interactions

```{r vowel-gam}
vowel_gam <- bam(
  f1 ~
    vow_voi +
    s(time_point, k = 6) +
    s(time_point, by = vow_voi, k = 6),
  data = vowels
)
summary(vowel_gam)
```

```{r vowel-gam-plot}
plot_smooths(vowel_gam, time_point, voicing, vowel, split = list(vow_voi = c("vowel", "voicing")))
```

