---
title: "An introduction to GAM(M)s"
author: "Stefano Coretta"
date: "12/07/2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
```

```{r data}
set.seed(8788)
x <- seq(0,100,1)
y <- ((runif(1, 10, 20) * x) / (runif(1, 0, 10) + x)) + rnorm(101, 0, 1)
x_1 <- seq(0, 100, 1)
y_1 <- ((runif(1, 10, 20) * x_1) / (runif(1, 0, 100) + x_1)) + rnorm(101, 0, 1)

sim_traj <- tibble(
  x = c(x, x_1),
  y = c(y, y_1),
  g = rep(c("a", "b"), each = 101)
)

sim_traj_a <- filter(sim_traj, g == "a")
```

## Linear models

Some (fairly) linear data...

```{r plot-data}
ggplot(sim_traj_a, aes(x, y)) +
  geom_point()
```

## Linear models

* A general formula: $y = \beta_0 + \beta_1x$
    * $y$ is the **outcome variable**
    * $x$ is the **predictor**
    * $\beta_0$ is the **intercept**
    * $\beta_1$ is the **slope**

* We know $x$ and $y$
    * we need to estimate $\beta_0$, $\beta_1$

* We can add more predictors
    * $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_nx_n$

## Linear models

* code in `R`

```{r lm, eval=FALSE, echo=TRUE}
lm(y ~ x, data = sim_traj)
```

* estimated intercept and slope

\centering

```{r plot-lm, out.height="150pt"}
ggplot(sim_traj_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Linear models

Is it linear?

```{r plot-lm-loess, message=FALSE}
ggplot(sim_traj_a, aes(x, y)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", colour = "red") +
  geom_smooth(linetype = "dashed")
```

## Linear models with non-linear data

How to account for non-linearity in a linear model?

* Use **higher-degree polynomials**
    * quadratic: $y = \beta_0 + \beta_1x + \beta_2x^2$
    * cubic: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3$
    * $n$th: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3 + ... + \beta_nx^n$

## Linear models with non-linear data

```{r}
ggplot(sim_traj_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
```

## Linear models with non-linear data

```{r}
ggplot(sim_traj_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3))
```

## Linear models with non-linear data

```{r}
ggplot(sim_traj_a, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10))
```
