---
title: "Introduction to GAM(M)S"
subtitle: "Generalised Additive (Mixed) Models"
author: "Dr Stefano Coretta"
date: "12-15 December 2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=7, fig.height=5, fig.retina=3,
  out.width = "60%", fig.align = "center",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
theme_set(theme_light())
library(mgcv)
library(tidygam)
library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
options(show.signif.stars = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(outfile = "slides/xaringan-themer.css")
```

```{r data, message=FALSE}
set.seed(8788)
x <- seq(0,100,1)
y <- ((runif(1, 10, 20) * x) / (runif(1, 0, 10) + x)) + rnorm(101, 0, 1)
x_1 <- seq(0, 100, 1)
y_1 <- ((runif(1, 10, 20) * x_1) / (runif(1, 0, 100) + x_1)) + rnorm(101, 0, 1)

sim_nl <- tibble(
  x = c(x, x_1),
  y = c(y, y_1),
  group = rep(c("a", "b"), each = 101)
)

sim_nl_a <- filter(sim_nl, group == "a")
```

class: center middle inverse

.f1[Time travel...]

---

layout: true

## Linear models

---

$$y = 3 + 2x$$

where $x = (2, 4, 5, 8, 10, 23, 36)$

---

```{r homework}
line <- tibble(
  x = c(2, 4, 5, 8, 10, 23, 36),
  y = 3 + 2 * x
)
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  labs(title = bquote(italic(y) == 3 + 2 * italic(x)))
```

---

```{r line}
ggplot(line, aes(x, y)) +
  geom_point(colour = "gray", size = 4) +
  geom_line(colour = "red") +
  labs(title = bquote(italic(y) == 3 + 2 * italic(x)))
```

---

* In research contexts, we have $x$ and $y$.

* For example, vowel duration and VOT, speech rate and pitch, etc.

---

```{r sampled}
set.seed(4321)
x <- 1:10
y <- (1 + 1.5 * x) + rnorm(10, 0, 1)

line <- tibble(
  x = x,
  y = y
)
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == "?" + "?" * italic(x)))
```

---

* The formula: $y = \beta_0 + \beta_1x$
    * $\beta_0$ is the **intercept**
    * $\beta_1$ is the **slope**

--

* We know $x$ and $y$.
    * We need to estimate $\beta_0$, $\beta_1$ = $\hat{\beta_0}, \hat{\beta_1}$.

--

* We can add more predictors:
    * $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_nx_n$

--

* `lm(y ~ x, data)` ($y$ as a function of $x$).

---

```{r lm-plot}
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote({italic(y) == beta[0] + beta[1] * italic(x)} == 1 + 1.5 * italic(x)))
```

---

```{r error}
m <- lm(y ~ x)
yhat <- m$fitted.values
diff <- y - yhat  
ggplot(line, aes(x, y)) +
  geom_segment(aes(x = x, xend = x, y = y, yend = yhat), colour = "red") +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + epsilon))
```

---

layout: false
layout: true

## LM with non-linear data

---

```{r plot-data}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  labs(title = "Some non-linear data")
```

---

```{r plot-lm}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Some non-linear data")
```

---

```{r error-2}
m <- lm(y ~ x, data = sim_nl_a)
yhat <- m$fitted.values
diff <- sim_nl_a$y - yhat  
ggplot(sim_nl_a, aes(x, y)) +
  geom_segment(aes(x = x, xend = x, y = y, yend = yhat), colour = "red") +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Some non-linear data")
```

---

How to account for non-linearity in a linear model?

--

* Use **higher-degree polynomials**
    * quadratic: $y = \beta_0 + \beta_1x + \beta_2x^2$
    * cubic: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3$
    * $n$th: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3 + ... + \beta_nx^n$

---

```{r 2-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2))
```

---

```{r 3-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3))
```

---

```{r 10-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10))
```

---

```{r 10-poly-2, warning=FALSE}
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10)) +
  ylim(1, 18)
```

---

```{r 10-poly-21, warning=FALSE}
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10)) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10)) +
  ylim(1, 18)
```

---

layout: false

## Generalised additive models

* **G**enrealised **A**dditive **M**odel**s** (GAMs)

* $y = f(x)$
    * $f(x)$ = some function of $x$ (or *smooth function*)

---

## Smooth terms

* LMs have **parametric terms**
    * $\beta_nx_n$
    * `x` in `R`
    * For linear effects.

--

* GAMs add (non-parametric) **smooth terms** (or simply smooths, also smoothers):
    * $f(x)$
    * `s(x)` in `R`
    * For non-linear effects.

--

* `library(mgcv); gam(y ~ s(x), data)`: $y$ as *some* function of $x$

---

## Smoothing splines, basis, basis functions

* Smooths in GAMs are **smoothing splines**.
    * Splines are defined piecewise with a set of polynomials.

--

* The set of polynomials is called a **basis**.
    * The basis is composed of **basis functions** (the polynomials).
    * The number of basis functions is called the **dimension of the basis** (`k` in mgcv).

--

* A spline is the sum of the products of each basis function and its coefficient.

---

## Basis functions and knots

```{r basis}
par(mfrow = c(1, 2))

simple <- gam(y ~ s(x, bs = "cr", k = 10), data = sim_nl_a)

X <- predict(
  simple,
  newdata = data.frame(x = sim_nl_a$x),
  type = "lpmatrix"
)[,2:10]

plot(sim_nl_a$x, X[, 1], type = "n", ylim = c(-2, 3), xlab = "x", ylab = "y")
for (i in 1:ncol(X)) {
  lines(sim_nl_a$x, X[, i], col = i, lw = 1)
}

X2 <- X %*% diag(coef(simple)[2:10])
plot(
  sim_nl_a$x,
  predict(simple, newdata = data.frame(x = sim_nl_a$x)) -
    coef(simple)[1],
  type = "n",
  ylim = c(-8, 3),
  xlab = "x", ylab = "y"
)
for (i in 1:ncol(X2)) {
  lines(sim_nl_a$x, X2[, i], col = i, lw = 1)
}
lines(
  sim_nl_a$x,
  predict(simple, newdata = data.frame(x = sim_nl_a$x)) -
    coef(simple)[1],
  lw = 3
)
```

---

## Smoothing parameter

* Wiggliness is related to **number of basis functions**.
    * more basis functions, more wiggliness (less smoothing).

--

* The **smoothing parameter** penalises wiggliness.
    * High values = less wiggliness (more smoothing).
    * This is estimated from the data.

---

## Smoothing splines

* There are **several kinds of splines**.
    * Each has its own basis functions.

* Most common splines:
    * *Thin plate regression splines*.
    * *Cubic regression splines*.

* For more info, run `?smooth.terms`.

---

layout: true

## A simple GAM

---

```{r simple-10, echo = TRUE}
simple <- gam(
  y ~ 
    # Cubic regression spline.
    # k is the dimension of the basis.
    s(x, bs = "cr", k = 10),
  data = sim_nl_a
)
```

---

```{r summary-10, echo = TRUE}
summary(simple)
```

---

```{r simple-10-plot}
predict_gam(simple, series = "x", length_out = 30) %>% plot()
```

---

```{r simple-5}
simple_5 <- gam(y ~ s(x, bs = "cr", k = 5), data = sim_nl_a)
predict_gam(simple_5, series = "x", length_out = 30) %>% plot()
```

---

layout: false
layout: true

## Comparing groups

---

```{r nl-plot}
ggplot(sim_nl, aes(x, y, colour = group)) +
  geom_point(size = 4)
```

---

```{r nl-smooth, message=FALSE}
ggplot(sim_nl, aes(x, y, colour = group)) +
  geom_point(size = 4, alpha = 0.3) +
  geom_smooth(se = FALSE, method = "gam")
```

---

* `by`-variables with ordered factors

```{r by-variables, echo=TRUE, eval=FALSE}
compare <- gam(
  y ~
    group +
    s(x, bs = "cr", k = 5) +
    s(x, bs = "cr", k = 5, by = group),
  data = sim_nl
)
```

---

To use `by`-variables with ordered factors:

* Change factor to an **ordered factor**.

* Change factor contrast to **treatment contrast** (`contr.treatment`).
    * The default in ordered factors is `contr.poly`, this won't work.
    
* Include the factor as a **parametric term**.

* Include a **reference smooth** and a **difference smooth** with the `by`-variable.

---

```{r ordered, echo=TRUE}
sim_nl <- sim_nl %>%
  mutate(group_o = ordered(group, levels = c("a", "b")))
contrasts(sim_nl$group_o) <- "contr.treatment"
```

---

```{r compare-gam, echo=TRUE}
compare <- gam(
  y ~
    # parametric term
    group_o +
    # reference smooth
    s(x, bs = "cr", k = 5) +
    # difference smooth
    s(x, bs = "cr", k = 5, by = group_o),
  data = sim_nl
)
```

---


```{r compare-summary}
summary(compare)
```

---

```{r plot-smooth, echo=TRUE}
library(tidygam)
predict_gam(compare, series = "x", length_out = 30) %>% plot(comparison = "group_o")
```

---

layout: false
class: center middle inverse

# HANDS-ON 1

---

layout: true

## Comparing across groups (interactions)

---

* Technically, GAMs **don't allow interactions**.

  * They are ADDITIVE (interactions require multiplication).

--

* We can get interaction-like comparisons by creating **factor interactions** and using them as `by`-variables.

---

Let's use the ACCH data frame from the hands-on.

```{r acch}
acch <- read_delim("data/ACCH.txt") %>%
  mutate(
    StimSlide.RT = StimSlide.RT/1000,
    RT_log = log(StimSlide.RT),
    ACC = ifelse(StimSlide.ACC == 0, "incorrect", "correct")
  )

acch
```

---

Now, let's create a factor interaction between `Group` and `Structure`.

We also need to make it into an ordered factor with treatment contrasts.

```{r acch-gr-st}
acch <- acch %>%
  mutate(
    Gr_St = as.ordered(interaction(Group, Structure))
  )
contrasts(acch$Gr_St) <- "contr.treatment"

head(acch$Gr_St)
```

---

```{r rt-gam}
rt_gam <- gam(
  RT_log ~ Gr_St + s(Trial) + s(Trial, by = Gr_St),
  data = acch
)
```

---

```{r rt-gam-summ}
summary(rt_gam)
```

---

```{r rt-gam-plot}
# library(tidygam)
predict_gam(rt_gam, series = "Trial", length_out = 50) %>%
  plot(comparison = "Gr_St")
```

---

```{r rt-gam-plot-2}
predict_gam(rt_gam, series = "Trial", length_out = 50,
            separate = list(Gr_St = c("Group", "Structure"))) %>%
  plot(comparison = "Group") +
  facet_grid(Structure ~ .)
```

---

```{r rt-gam-diff}
get_difference(
  rt_gam, "Trial", list(Gr_St = c("bilingual.Object", "monolingual.Object"))
) %>%
  plot()
```

---

```{r rt-gam-diff-2}
get_difference(
  rt_gam, "Trial", list(Gr_St = c("bilingual.Object", "bilingual.Subject"))
) %>%
  plot()
```

---

```{r rt-gam-diff-3}
get_difference(
  rt_gam, "Trial", list(Gr_St = c("monolingual.Object", "monolingual.Subject"))
) %>%
  plot()
```

---

layout: false
layout: true

## Tensor product smooths and interactions

---

|                 | LM      | GAM                    |
|-----------------|---------|------------------------|
| Num × cat       | `num:cat` | `s(num, by = "cat")`     |
| Num × cat × cat | `num:cat:cat` | `s(num, by = "cat_cat")` |

---

|                 | LM      | GAM                    |
|-----------------|---------|------------------------|
| Num × cat       | `num:cat` | `s(num, by = cat)`     |
| Num × cat × cat | `num:cat:cat` | `s(num, by = cat_cat)` |
| Num × num       | `num:num` | `te(num, num)`           |
| Num × num × cat | `num:num:cat` | `te(num, num, by = cat)` |

---

* **Tensor product smooths** are smooths that create interactions between two numeric predictors.

* Tensor product smooths are specified with `te()` or `ti()`.

  * `ti(x) + ti(y) + ti(x, y)`.
  * `te(x, y)`.

--

Note that if the two numeric predictors are on the same scale (isotropic), you can use `s()` instead of `te/ti()`. Coordinates are such type of predictors.

---

```{r vowels, echo=TRUE}
vowels <- read_csv("data/vowels.csv")
vowels
```


---

```{r f0-gam, echo=TRUE}
f0_gam <- gam(
  f0 ~ 
    ti(time_point) + ti(duration) + ti(time_point, duration),
  data = vowels
)
```

---

```{r f0-gam-plot, echo=TRUE}
predict_gam(f0_gam) %>%
  plot(series = "time_point", comparison = "duration")
```

---

```{r f0-gam-plot-2, echo=TRUE}
predict_gam(f0_gam, values = list(duration = c(50, 100, 150, 200))) %>%
  plot(series = "time_point", comparison = "duration")
```

---

```{r f0-gam-plot-3, echo=TRUE}
predict_gam(f0_gam) %>%
  plot(
    series = c("time_point", "duration"),
    raster_interp = TRUE
  )
```

---

layout: false
class: center middle inverse

# HANDS-ON 2

---

layout: true

## Random effects with factor smooths

---

* Only **fixed effects** so far...

* **G**eneralised **A**dditive **M**ixed **M**odel (GAMM)

* Include a **random smooth** term with the **factor smooth interaction** as basis.

---

* **Factor smooth interaction**:
    * `bs = "fs"`.
    * A smooth is fitted at each level of a factor.

* The random effect variable *needs to be a factor*.

* `s(Trial, Subject, bs = "fs", m = 1)`

---

```{r acch-str-o}
acch <- acch %>%
  mutate(Structure_o = as.ordered(Structure))

contrasts(acch$Structure_o) <- "contr.treatment"
```

---

```{r rt-gam-2}
rt_gam_2 <- gam(
  RT_log ~
    Gr_St + s(Trial) + s(Trial, by = Gr_St) +
    s(Trial, Subject, bs = "fs", m = 1) +
    s(Trial, Subject, by = Structure_o, bs = "fs", m = 1),
  data = acch
)
```

---

```{r rt-gam-2-sum}
summary(rt_gam_2)
```


---

```{r rt-gam-2-plot}
predict_gam(
  rt_gam_2,
  separate = list(Gr_St = c("Group", "Structure")),
  exclude_terms = c("s(Trial,Subject)", "s(Trial,Subject):Structure_oSubject")
) %>%
  select(-Structure_o) %>%
  distinct() %>%
  plot(series = "Trial", comparison = "Structure") +
  facet_grid(~ Group)
```

---

```{r rt-gam-2-diff}
get_difference(
  rt_gam_2, "Trial", list(Gr_St = c("bilingual.Object", "bilingual.Subject"))
) %>%
  plot()
```


---

```{r rt-gam-2-diff-2}
get_difference(
  rt_gam_2, "Trial", list(Gr_St = c("monolingual.Object", "monolingual.Subject"))
) %>%
  plot()
```

---

layout: false

## GAMs with binomial/Bernoulli families

---

```{r acc-gam-2}
acc_gam_2 <- gam(
  StimSlide.ACC ~
    Gr_St + s(Trial) + s(Trial, by = Gr_St) +
    s(Trial, Subject, bs = "fs", m = 1),
  data = acch,
  family = binomial()
)
```

---

```{r acc-gam-2-sum}
summary(acc_gam_2)
```


---

```{r acc-gam-2-plot}
predict_gam(
  acc_gam_2,
  separate = list(Gr_St = c("Group", "Structure")),
  tran_fun = plogis,
  exclude_terms = "s(Trial,Subject)",
  length_out = 30
) %>%
  plot(series = "Trial", comparison = "Structure") +
  facet_grid(~ Group)
```

