---
title: "Introduction to GAM(M)S"
subtitle: "Generalised Additive (Mixed) Models"
author: "Dr Stefano Coretta"
date: "12 December 2022"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=7, fig.height=5, fig.retina=3,
  out.width = "60%", fig.align = "center",
  cache = FALSE,
  echo = FALSE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
library(tidyverse)
theme_set(theme_light())
library(mgcv)
library(tidygam)
library(xaringanExtra)
use_xaringan_extra(c("panelset", "tachyons", "freezeframe"))
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light()
```

```{r data, message=FALSE}
set.seed(8788)
x <- seq(0,100,1)
y <- ((runif(1, 10, 20) * x) / (runif(1, 0, 10) + x)) + rnorm(101, 0, 1)
x_1 <- seq(0, 100, 1)
y_1 <- ((runif(1, 10, 20) * x_1) / (runif(1, 0, 100) + x_1)) + rnorm(101, 0, 1)

sim_nl <- tibble(
  x = c(x, x_1),
  y = c(y, y_1),
  group = rep(c("a", "b"), each = 101)
)

sim_nl_a <- filter(sim_nl, group == "a")
```

class: center middle inverse

.f1[Time travel...]

---

layout: true

## Linear models

---

$$y = 3 + 2x$$

where $x = (2, 4, 5, 8, 10, 23, 36)$

---

```{r homework}
line <- tibble(
  x = c(2, 4, 5, 8, 10, 23, 36),
  y = 3 + 2 * x
)
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  labs(title = bquote(italic(y) == 3 + 2 * italic(x)))
```

---

```{r line}
ggplot(line, aes(x, y)) +
  geom_point(colour = "gray", size = 4) +
  geom_line(colour = "red") +
  labs(title = bquote(italic(y) == 3 + 2 * italic(x)))
```

---

* In research contexts, we have $x$ and $y$.

* For example, vowel duration and VOT, speech rate and pitch, etc.

---

```{r sampled}
set.seed(4321)
x <- 1:10
y <- (1 + 1.5 * x) + rnorm(10, 0, 1)

line <- tibble(
  x = x,
  y = y
)
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == "?" + "?" * italic(x)))
```

---

* The formula: $y = \beta_0 + \beta_1x$
    * $\beta_0$ is the **intercept**
    * $\beta_1$ is the **slope**

--

* We know $x$ and $y$.
    * We need to estimate $\beta_0$, $\beta_1$ = $\hat{\beta_0}, \hat{\beta_1}$.

--

* We can add more predictors:
    * $y = \beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_nx_n$

--

* `lm(y ~ x, data)` ($y$ as a function of $x$).

---

```{r lm-plot}
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote({italic(y) == beta[0] + beta[1] * italic(x)} == 1 + 1.5 * italic(x)))
```

---

```{r error}
m <- lm(y ~ x)
yhat <- m$fitted.values
diff <- y - yhat  
ggplot(line, aes(x, y)) +
  geom_segment(aes(x = x, xend = x, y = y, yend = yhat), colour = "red") +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + epsilon))
```

---

layout: false
layout: true

## LM with non-linear data

---

```{r plot-data}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  labs(title = "Some non-linear data")
```

---

```{r plot-lm}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Some non-linear data")
```

---

```{r error-2}
m <- lm(y ~ x, data = sim_nl_a)
yhat <- m$fitted.values
diff <- sim_nl_a$y - yhat  
ggplot(sim_nl_a, aes(x, y)) +
  geom_segment(aes(x = x, xend = x, y = y, yend = yhat), colour = "red") +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Some non-linear data")
```

---

How to account for non-linearity in a linear model?

--

* Use **higher-degree polynomials**
    * quadratic: $y = \beta_0 + \beta_1x + \beta_2x^2$
    * cubic: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3$
    * $n$th: $y = \beta_0 + \beta_1x + \beta_2x^2 + \beta_3x^3 + ... + \beta_nx^n$

---

```{r 2-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2))
```

---

```{r 3-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3))
```

---

```{r 10-poly}
ggplot(sim_nl_a, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10)) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10))
```

---

```{r 10-poly-2, warning=FALSE}
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10)) +
  ylim(1, 18)
```

---

```{r 10-poly-21, warning=FALSE}
ggplot(line, aes(x, y)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x + I(x^2) + I(x^3) + I(x^4) + I(x^5) + I(x^6) + I(x^7) + I(x^8) + I(x^9) + I(x^10)) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = bquote(italic(y) == beta[0] + beta[1] * italic(x) + beta[2] * italic(x) ^ 2 + beta[3] * italic(x) ^ 3 + beta[4] * italic(x) ^ 4 + beta[5] * italic(x) ^ 5 + beta[6] * italic(x) ^ 6 + beta[7] * italic(x) ^ 7 + beta[8] * italic(x) ^ 8 + beta[9] * italic(x) ^ 9 + beta[10] * italic(x) ^ 10)) +
  ylim(1, 18)
```

---

layout: false

## Generalised additive models

* **G**enrealised **A**dditive **M**odel**s** (GAMs)

* $y = f(x)$
    * $f(x)$ = some function of $x$ (or *smooth function*)

---

## Smooth terms

* LMs have **parametric terms**
    * $\beta_nx_n$
    * `x` in `R`
    * For linear effects.

--

* GAMs add (non-parametric) **smooth terms** (or simply smooths, also smoothers):
    * $f(x)$
    * `s(x)` in `R`
    * For non-linear effects.

--

* `library(mgcv); gam(y ~ s(x), data)`: $y$ as *some* function of $x$

---

## Smoothing splines, basis, basis functions

* Smooths in GAMs are **smoothing splines**.
    * Splines are defined piecewise with a set of polynomials.

--

* The set of polynomials is called a **basis**.
    * The basis is composed of **basis functions** (the polynomials).
    * The number of basis functions is called the **dimension of the basis** (`k` in mgcv).

--

* A spline is the sum of the products of each basis function and its coefficient.

---

## Basis functions and knots

```{r basis}
par(mfrow = c(1, 2))

simple <- gam(y ~ s(x, bs = "cr", k = 10), data = sim_nl_a)

X <- predict(
  simple,
  newdata = data.frame(x = sim_nl_a$x),
  type = "lpmatrix"
)[,2:10]

plot(sim_nl_a$x, X[, 1], type = "n", ylim = c(-2, 3), xlab = "x", ylab = "y")
for (i in 1:ncol(X)) {
  lines(sim_nl_a$x, X[, i], col = i, lw = 1)
}

X2 <- X %*% diag(coef(simple)[2:10])
plot(
  sim_nl_a$x,
  predict(simple, newdata = data.frame(x = sim_nl_a$x)) -
    coef(simple)[1],
  type = "n",
  ylim = c(-8, 3),
  xlab = "x", ylab = "y"
)
for (i in 1:ncol(X2)) {
  lines(sim_nl_a$x, X2[, i], col = i, lw = 1)
}
lines(
  sim_nl_a$x,
  predict(simple, newdata = data.frame(x = sim_nl_a$x)) -
    coef(simple)[1],
  lw = 3
)
```

---

## Smoothing parameter

* Wiggliness is related to **number of basis functions**.
    * more basis functions, more wiggliness (less smoothing).

--

* The **smoothing parameter** penalises wiggliness.
    * High values = less wiggliness (more smoothing).
    * This is estimated from the data.

---

## Smoothing splines

* There are **several kinds of splines**.
    * Each has its own basis functions.

* Most common splines:
    * *Thin plate regression splines*.
    * *Cubic regression splines*.

* For more info, run `?smooth.terms`.

---

layout: true

## A simple GAM

---

```{r simple-10, echo = TRUE}
simple <- gam(
  y ~ 
    # Cubic regression spline.
    # k is the dimension of the basis.
    s(x, bs = "cr", k = 10),
  data = sim_nl_a
)
```

---

```{r summary-10, echo = TRUE}
summary(simple)
```

---

```{r simple-10-plot}
predict_gam(simple, series = "x", length_out = 30) %>% plot()
```

---

```{r simple-5}
simple_5 <- gam(y ~ s(x, bs = "cr", k = 5), data = sim_nl_a)
predict_gam(simple_5, series = "x", length_out = 30) %>% plot()
```
