---
title: "Practical 2"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(mgcv)
library(itsadug)
library(tidymv)

load("./www/practical-2.Rdata")
```

## Dynamic data: F0 trajectory

In `Practical 1` we looked at F0 as a function of speech rate. In this practical, we will instead model F0 trajectories. F0 trajectories are a type of dynamic data, in particular they are time series data.

### Revision

Let's start with a GAM that models the change in F0 through time. We just need a smooth term with `time_point`. I filled in some code in the following exercise. Add the missing bits and then run the code!

```{r gam-1, exercise=TRUE}
# Include the outcome variable and a smooth term with time_point

f0_gam <- bam(
  ... ~
    s(...),
  data = vowels
)

summary(f0_gam)
```

```{r gam-1-solution}
# Include the outcome variable and a smooth term with time_point

f0_gam <- bam(
  f0 ~
    s(time_point, k = 6),
  data = vowels
)

summary(f0_gam)
```

And now let's plot the smooth with `plot_smooths()`.

```{r gam-plot, exercise=TRUE, exercise.lines=1}

```

### Effect of vowel and voicing

Now we want to check the effect of the indentity of the vowel (`vowel`) and the voicing of the consonant following the vowel (`voicing`). To do so we use smooths with `by`-variables.

Remember that factors in a `by`-variable need to be ordered factors with treatment contrasts. I have already changed the factors in `vowels` for you (if you don't remember how to do this, check Part I on the slides.)

Let's fit a model with difference smooths for `vowel` and `voicing`.

```{r gam-2, exercise=TRUE}
f0_gam_2 <- bam(
  f0 ~
    ... +
    ... +
    s(time_point, k = 6) +
    s(time_point, k = 6, by = ...) +
    s(time_point, k = 6, by = ...),
  data = vowels
)

summary(f0_gam_2)
```

## Random effects

Our `vowel` dataset has measurements of F0 from several speakers and several words. It is important then to include `speaker` and `word` as random effects.

Random effects can be included as random smooths (remember that they need to be factors).

```{r quiz-1}
quiz(
  question("What is the basis for specifying random smooths?",
    answer("factor smooth (`bs`)"),
    answer("ordered factor (`by`)"),
    answer("factor smooth interaction (`fs`)", correct = TRUE),
    answer("random effect (`re`)", message = "`re` is used for classic random intercepts and slopes.")
  )
)
```

### Random smooth

Let's add random smooths for `speaker` and `word` and run the code (It might take a few seconds to run).

```{r gam-3, exercise=TRUE}
f0_gam_3 <- bam(
  f0 ~
    vowel +
    voicing +
    s(time_point, k = 6) +
    s(time_point, k = 6, by = vowel) +
    s(time_point, k = 6, by = voicing) +
    # your random smooths
    s(time_point, ..., ..., ...) +
    s(time_point, ..., ..., ...),
  data = vowels
)

summary(...)
```

<div id="gam-3-hint">
Did you remember to chage `k` in the random smooths?
</div>

```{r gam-3-solution}
f0_gam_3 <- bam(
  f0 ~
    vowel +
    voicing +
    s(time_point, k = 6) +
    s(time_point, k = 6, by = vowel) +
    s(time_point, k = 6, by = voicing) +
    # your random smooths
    s(time_point, speaker, bs = "fs", k = 6) +
    s(time_point, word, bs = "fs", k = 6),
  data = vowels
)

summary(f0_gam_3)
```

### Plot the model

Now we can plot the model with `plot_smooths()`. Remember that by default this command removes the random smooths from the estimates, so that the average estimate smooth is plotted.

```{r f0-gam-3-plot, echo=TRUE}
plot_smooths(f0_gam_3, time_point, comparison = voicing, facet_terms = vowel)
```

## `by`-interactions

In the previous models we were including `vowel` and `voicing` as factors in `by`-variable smooths. In this way, we are modelling their effect separately. When plotting the estimated smooths, the individual effects `vowel` and `voicing` are added up to each other (hence the 'additive' part in 'generalised additive models').

Most of the times we expect that, for example, different vowels behave differently with respect to other speech properties: what we are looking for is classic interactions.

Classic interactions in GAMs (like the ones you add in `lmer` models with the colon `:` notation) can be included using `by`-variables. The factor in the `by`-variable will be an interaction factor.

You can create an interaction factor using `interaction()` with the terms that constitute the interaction (in our case, vowel and voicing). The factor needs to be an ordered factor with treatment contrasts, like any other `by`-variable factor. I have already prepared the interaction factor for you with the following code.

```{r inter, echo=TRUE, eval=FALSE}
vowels <- vowels %>%
  mutate(
    vow_voi = interaction(vowel, voicing),
    vow_voi = as.ordered(vow_voi)
  )

contrasts(vowels$vow_voi) <- "contr.treatment"
```

### Include a `by`-interaction in the model

Now we can construct a model with an interaction between `vowel` and `voicing`.
By including a `by`-interaction smooth, separate smooths will be fitted for each combination of `vowel` and `voicing`, allowing different effects of voicing on F0 by vowel.

Add the necessary terms in the following exercise (it will take a few seconds to run).

```{r f0-gam-4, exercise=TRUE}
f0_gam_4 <- bam(
  f0 ~
    ...
    s(time_point, k = 6) +
    ...
    s(time_point, speaker, bs = "fs", k = 6) +
    s(time_point, word, bs = "fs", k = 6),
  data = vowels
)

summary(f0_gam_4)
```

<div id="f0-gam-4-hint">
Remember to include both a parametric term and a difference smooth.
</div>

### And now let's plot!

Now we can plot the model with a `vowel:voicing` interaction. Since we are using a `by`-interaction, we need to specify the `split` argument in `plot_smooths()` (I added it for you), which splits the interaction back into the `vowel` and `voicing` factors.

```{r f0-gam-4-plot, exercise=TRUE}
plot_smooths(
  f0_gam_4,
  time_series = ...,
  comparison = ...,
  facet_terms = ...,
  split = list(vow_voi = c("vowel", "voicing"))
)
```

```{r f0-gam-4-plot-solution}
plot_smooths(f0_gam_4, time_point, comparison = voicing, facet_terms = vowel, split = list(vow_voi = c("vowel", "voicing")))
```
