---
title: "Practical 1"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
theme_set(theme_minimal())
library(mgcv)
library(itsadug)
library(tidymv)

# Simulate data
set.seed(2018)
n <- 500
x <- 1:n
y <- 3 * log(x) + 10 + rnorm(n, 0, 1)
x_1 <- 1:n
y_1 <- 3 * log(x) + 4 + rnorm(n, 0, 1)

sim_data <- tibble(
  speech_rate = c(x, x_1),
  F0 = c(y, y_1),
  focus = rep(c("focus", "non-focus"), each = n)
) %>%
  mutate(
    focus_o = ordered(focus, levels = c("non-focus", "focus"))
  )
contrasts(sim_data$focus_o) <- "contr.treatment"
```


## Fit a simple GAM

We will start by fitting a simple GAM to a big(ish) simulated dataset (*n* = 1000), `sim_data`. This dummy dataset includes measurements of **speech rate** and **fundamental frequency** (F0).[^1] We want to know what is the relationship between speech rate and F0. The fictional experiment also included two focus conditions: *focus* and *non-focus*. So we are particularly interested in weather the condition *focus* has an effect on the relationship between speech rate and F0.

In this practical, we will build our models incrementally, for pedagogical purpuses. In real life applications, of course, you will prepare the data for fitting and start from a model with all of the terms you think are necessary.


### The data

Let's first have a look at how the dataset looks like. Run the following code.

```{r sim-data, exercise=TRUE, exercise.lines=1}
sim_data
```

Now let's plot the data.

```{r plot-data, exercise=TRUE}
sim_data %>%
  ggplot(aes(speech_rate, F0)) +
  geom_point(alpha = 0.7, colour = "orchid4")
```

There is a clear non-linear relationship between speech rate and F0: when speech rate increases, F0 increases too, but at faster speech rates the increase in F0 is smaller. Also, we can see a separation between two groups in the data, but we will deal with this later.

We can model this non-linear increase in F0 using a GAM.

### Fit a GAM

Let's now fit a GAM to F0. We will use the package `mgcv` and the function `gam()` from that package to fit a GAM.

Refresher:

* `s()` defines a smooth (a curve) over a variable, in this case `speech_rate`
* `bs` is used to select the type of *basis*
    * `tp` is a thin plate regression spline (and it is also the default)
* `k` is the number of *knots* (= number of basis functions + 1)

```{r F0-gam, exercise = TRUE}
library(mgcv)

F0_gam <- gam(
  # the outcome variable: F0
  F0 ~
    # a smooth for speech rate
    s(speech_rate, bs = "tp", k = 10),
  # the data
  data = sim_data
)
summary(F0_gam)
```

```{r quiz-1}
quiz(
  question("Is there a non-linear correlation between `speech rate` and `F0`? (Hint: check the EDF value of `s(speech_rate)` and the *p*-value)",
    answer("yes", correct = TRUE, message = "When `edf > 1`, the smooth is non-flat."),
    answer("no", message = "When `edf = 1`, the smooth is flat.")
  )
)
```

### Plot a single smooth

```{r plot-gam-setup}
F0_gam <- gam(
  # the outcome variable: F0
  F0 ~
    # a smooth for speech rate
    s(speech_rate, bs = "tp", k = 10),
  # the data
  data = sim_data
)
```

Let's now plot the model with `plot_smooths()` from `tidymv`!

```{r plot-gam, exercise=TRUE, exercise.lines=1}
plot_smooths(F0_gam, speech_rate)
```

If you want, try different basis and different values of k and plot the result in the following exercise (in particular, try decreasing `k`).

```{r try, exercise=TRUE}
F0_gam <- gam(
  # the outcome variable: F0
  F0 ~
    # a smooth for speech rate
    s(speech_rate, bs = "tp", k = 10),
  # the data
  data = sim_data
)

plot_smooths(F0_gam, speech_rate)
```

[^1]: If you want to see the code for generating the data, look at the file `practical-1.Rmd`, in the `setup` chunk file.






## Compare focus vs. non-focus

### Prepare ordered factors

What we are really interested in is to compare the effect of **speech rate** on **F0** in the two focus conditions: *focus* and *non-focus*. We can do this by setting up a model using a `by` variable with an ordered factor. Remember that to use a `by` variable we need to create an ordered factor and change the contrasts to treatment. I did this for you already, using the following code.

```{r ordered, echo=TRUE}
sim_data <- sim_data %>%
  mutate(
    focus_o = ordered(focus, levels = c("non-focus", "focus"))
  )
contrasts(sim_data$focus_o) <- "contr.treatment"
```

### Fit with a `by` variable

We can now fit a GAM using a smooth with a `by` variable. To do so, we need to add `focus` as a parametric term, include a smooth without the `by` variable (the reference smooth) and one with it (the difference smooth).

```{r focus-gam, exercise=TRUE}
focus_gam <- gam(
  # the outcome variable: F0
  F0 ~
    # the parametric term
    focus_o +
    # the reference smooth
    s(speech_rate, bs = "tp", k = 10) +
    # the difference smooth
    s(speech_rate, by = focus_o, bs = "tp", k = 10),
  # the data
  data = sim_data,
  method = "ML"
)
summary(focus_gam)
```

```{r focus-setup}
focus_gam <- gam(
  # the outcome variable: F0
  F0 ~
    # the parametric term
    focus_o +
    # the reference smooth
    s(speech_rate, bs = "tp", k = 10) +
    # the difference smooth
    s(speech_rate, by = focus_o, bs = "tp", k = 10),
  # the data
  data = sim_data,
  method = "ML"
)
```

And finally let's plot the estimated smooths with `tidymv::plot_smooths()`.

```{r plot-focus, exercise=TRUE, exepriment.lines=1}
plot_smooths(focus_gam, speech_rate, comparison = focus_o)
```

### Significance testing

To test for significance, we fit a null model and compare this model with the full model using `compareML()`. The null model must exclude *both* the parametric term and the difference smooth.

```{r compare-focus-gam, exercise=TRUE, exercise.setup="focus-setup"}
focus_gam_null <- gam(
  F0 ~
    s(speech_rate, bs = "tp", k = 10),
  data = sim_data,
  method = "ML"
)

compareML(focus_gam_null, focus_gam)
```

Is there a significant difference between focus and non-focus?
