---
title: "Hands-on 1"
author: "Stefano Coretta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
theme_set(theme_light())
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
library(mgcv)
options(show.signif.stars = FALSE)
library(tidygam)
```

## Read data

For this hands-on, we will use data kindly offered by George Pontikas, Ian Cunnings, and Theodoros Marinis (<https://osf.io/4w693/>).

Let's read it in and change a few things to make things easier later.

```{r acch}
acch <- read_delim("data/ACCH.txt") %>%
  mutate(
    StimSlide.RT = StimSlide.RT/1000,
    RT_log = log(StimSlide.RT),
    ACC = ifelse(StimSlide.ACC == 0, "incorrect", "correct")
  )

acch
```

For now we can focus on the following columns:

- `Subject`: the subject ID.
- `Group`: whether the subject is `monolingual` or `bilingual`.
- `Trial`: the number of the trial.
- `StimSlide.RT`: the reaction times in milliseconds.

## Fit a simple GAM

In this practical, we will build our models incrementally, for pedagogical purposes. In real-life applications, of course, you will prepare the data for fitting and start from a model with all of the terms you think are necessary.

Now let's plot the data.

```{r acch-plot}
acch %>%
  ggplot(aes(Trial, RT_log)) +
  geom_smooth(method = "gam", formula = y ~ s(x))
```

There is a clear non-linear relationship between trial number and logged reaction times: RTs decrease in the first fourth of the trials, they plateau in the second fourth, and increase again through out the second half of the trials.

We are discarding for now differences between groups and other variables.

### Fit the GAM

Let's now fit a GAM to logged RTs. We use the `gam()` function from the package `mgcv` to fit a GAM.

Quick refresher:

* `s()` defines a smooth over a variable, in this case `Trial`.
* `bs` is used to select the type of *basis*
    * `tp` is a thin plate regression spline (and it is also the default).
* `k` is dimension of the basis.

```{r rt-gam}
library(mgcv)

rt_gam <- gam(
  # The outcome variable: RT_log
  RT_log ~
    # A smooth for trial number
    s(Trial, bs = "tp", k = 10),
  # the data
  data = acch
)

summary(rt_gam)
```

The model summary is made of different parts:

- **Family and Link function**
  - They report the distribution family and link used in the model. Here, that is Gaussian with an identity link.
- **Formula**
  - The model formula.
  
- **Parametric coefficients**
  - The estimated coefficients of the parametric (linear) terms. Here, just the intercept.
  - The parametric intercept can be thought of as the overall mean across all trials.

- **Approximate significance of smooth terms**
  - Here, the statistical significance of smooth terms is reported. Note that for smooth terms you don't get a summary like you do for parametric terms.
  - `edf` are the Estimated Degrees of Freedom. This measure is related to the number of basis functions selected for that particular term and also takes into account the smoothing penalty.
  - An EDF of 1 means that the model estimated the effect of that smooth to be linear (a straight line) rather than non-linear.
  - An EDF above 1 mean that the model estimated the effect to be non-linear.
  - The *p*-value indicates if the EDF is significantly different from 1 (rather than 0).
  - *p*-values below 0.05 indicate that the estimated effect is significantly different from a linear effect.
  
- **Further model info**
  - R^2 and deviance explained are two measures of model goodness.
    - Deviance explained is more robust than R^2.
  - GCV and scale are specific model parameters.
    - The GCV score is the minimised generalised cross-validation (GCV) score of the GAM fitted. GCV is used for smoothness selection.
    - The scale is the squared residual standard error for Gaussian models.
    - For a more technical explanation, see <https://stats.stackexchange.com/a/191235/128897>.
  - N is the sample size.

In summary, the simple model indicates that the overall mean RT is 1.6 log RT and that the effect of trial number on RT is non-linear.

To know more about the predicted values, we need to plot them.

### Get model predictions and plot them

Model predictions can be conveniently extracted with the tidygam package (under development here: <https://github.com/stefanocoretta/tidygam>).

You can install the development version with `remotes::install_github("stefanocoretta/tidygam@devel")` (you also need the remotes package).

The `predict_gam()` function extracts predictions based on all the predictors in the model (all levels are included for categorical predictors and a range of values are selected for numeric predictors). It returns a tibble.

The standard error and lower and upper Confidence Interval boundaries are also returned.

```{r rt-gam-preds}
library(tidygam)

rt_gam_preds <- predict_gam(rt_gam)
rt_gam_preds
```

We can easily plot the model predictions with the `plot()` function.

You need to specify the series (*x*-axis) for plotting. Here, that is `Trial`.

```{r rt-gam-plot}
rt_gam_preds %>% plot(series = "Trial")
```

Note that the Confidence Intervals are Bayesian Credible Intervals. This means that they can be interpreted as such: A 95% Bayesian Credible Interval indicates the range within which we can be 95% confident the real value lies. (This is different from a frequentist Confidence Interval, which is very often misinterpreted as a Bayesian Credible Interval.)

You can adjust the resolution (number of points) of the predicted smooth by specifying the `length_out` argument in `predict_gam()`.
Higher `length_out` values make the smooth look smoother. Note that this is independent of the smooth's smoothness parameter, and just affects plotting.

```{r rt-gam-plot-2}
rt_gam_preds_2 <- predict_gam(rt_gam, length_out = 20) %>% plot(series = "Trial")
rt_gam_preds_2
```

Plotting the output of `predict_gam()` returns a ggplot object, which you can further modify using ggplot2 functions:

```{r rt-gam-plot-3}
rt_gam_preds_2 +
  labs(title = "Predicted log RTs by trial")
```

For more control, you can simply plot the predicted values using `ggplot()` as you would do with any other tibble.

```{r rt-gam-plot-4}
rt_gam_preds %>%
  ggplot(aes(Trial, RT_log)) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci), alpha = 0.2, fill = "#1B9E77") +
  geom_line(colour = "#1B9E77", linewidth = 1.5)
```

### Reporting

TODO

## Compare monolinguals vs bilinguals

### Prepare ordered factors

Let's compare monolingual vs bilingual participants. We can do this by setting up a model using a `by`-variable with `Group` as an ordered factor. Remember that to use a `by`-variable we need to create an ordered factor and change the contrasts to treatment.

We can achieve that with the following code.

```{r ordered}
acch <- acch %>%
  mutate(
    Group_o = ordered(Group, levels = c("monolingual", "bilingual"))
  )
contrasts(acch$Group_o) <- "contr.treatment"
```

Check the `acch` tibble in the Environment tab of the top-right panel in RStudio. If you click on the blue arrow button next to the tibble name, you'll see the new `Group_o` column: it is an `Ord.factor` and has `contr.treatment` as the `contrasts` attribute.

### Use the `by`-variable

We can now fit a GAM using a smooth with a `by`-variable.

To do so, we need to add `Group_o` as a parametric term, include a smooth without the `by`-variable (the reference smooth) and one with it (the difference smooth).

```{r rt-gam-2}
rt_gam_2 <- gam(
  # the outcome variable: RT_log
  RT_log ~
    # the parametric term
    Group_o +
    # the reference smooth
    s(Trial, bs = "cr", k = 10) +
    # the difference smooth
    s(Trial, by = Group_o, bs = "cr", k = 10),
  # the data
  data = acch
)
summary(rt_gam_2)
```

The summary structure is the same as before, but now we also have a parametric effect for `Group_o = bilingual` (think of this as the mean difference between log RTs of bilingual vs monolingual subjects).

The parametric term `Group_obilingual` indicates that on average log RT are 0.03 logs longer in the bilingual group compared to the monolingual group. This is a significant effect according to the reported *p*-value (*p* = 0.0104).

Parametric terms like `Intercept` and `Group_obilingual` basically model the **height** of the predicted smooth (for the monolingual and bilingual participants).

The **shape** of the smooth is modelled by the smooth terms.

Moving onto the smooth terms, now we have two terms:

- `s(Trial)` is our **reference smooth** (i.e. the smooth for monolingual subjects, since `monolingual` is the reference level of `Group_o`).

- `s(Trial):Group_obilingual` is the **difference smooth**. This corresponds to the predicted difference (smooth) between the `bilingual` and the `monolingual` smooths.

When the difference smooth EDF are significantly different from 1, then the model indicates that the shape of the smooth for that level is significantly different from the shape of the reference level, when controlling for the difference in height (which is modelled by the parametric effects).

Since the approximate *p*-value for `s(Trial):Group_obilingual` is 0.3749, above the 0.05 threshold, the model indicates that the *shape* of the smooth for bilingual participants is not significantly different from that of monolingual participants.

In other words, while the average log RTs are different in mono vs bilingual subjects, the effect of trial is suggested to be similar if not identical in both groups.

### Plot the model predictions

Let's plot the estimated smooths to see what all that means in practice.

We can use the `comparison` argument in `plot()` to colour the two levels in `Group_o`.

```{r rt-gam-2-plot}
predict_gam(rt_gam_2, length_out = 30, series = "Trial") %>%
  plot(comparison = "Group_o")
```

### Reporting

TODO

## Exercise 1

Now try this yourself! This time run a GAM with `Structure` as a `by`-variable instead of `Group_o`.

Remember that you first have to mutate `Structure` into an **ordered factor** and change the **contrasts to treatment**.

Do this here:

```{r structure-o}
acch <- acch %>%
  mutate(
    
  )
```

Now fit a GAM and print the summary.

```{r rt-gam-3}

```

What can you tell about the effect of `Structure` based on the summary?

Plot the predicted smooths to see if your intuitions match what the plot suggests.

```{r rt-gam-3-plot}

```

