---
title: "Hands-on 4"
author: "Stefano Coretta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
theme_set(theme_light())
library(ggforce)
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
library(mgcv)
options(show.signif.stars = FALSE)
library(tidygam)
```

## Massive Auditory Lexical Decision

In this exercise you will use a subset of the data in the Massive Auditory Lexical Decision data set (https://doi.org/10.7939/r3-kh0r-r116).

```{r mald}
load("data/mald_1_1.rda")
```

The subset contains data from 50 subjects, 100 trials each:

- `Subject`: the subject's ID.
- `Item`: the target word.
- `IsWord`: whether the target word is a real `"TRUE"` or a nonce word `"FALSE"`.
- `PhonLev`: mean phonological Levenshtein distance (higher value means the word is more unique, i.e. a higher mean number of changes have to be applied to generate the other words).
- `RT`: lexical decision reaction times in ms.
- `ACC`: accuracy as `incorrect` or `correct`.
- `RT_log`: logged reaction times.

Let's model accuracy `ACC` with a binomial/Bernoulli GAMM as a function of `IsWord`, `PhonLev` and `RT_log`. We also want to include random smooths for `Subject`.

First, create the necessary factor interactions.

```{r}

```

Now fit the GAMM.

```{r}

```

Check the model with `gam.check()`. Is the number of basis functions enough?

```{r}

```

If everything is ok, get the predictions of the model and plot it.

```{r}

```

What can you tell about the data based on the model summary and plots?


## Infant gestures

This exercise uses data from infants of Bengali, Chinese and English cultural backgrounds (<https://www.doi.org/10.1111/cdev.13406>).

```{r gestures}
load("data/gestures.rda")
```

- `dyad`: mother/infant dyad ID.
- `background`: infant background (`Bengali`, `Chinese`, `English`).
- `month`: age of infant in months (`10`, `11`, `12`).
- `task`: task type (`five`, `tp1`, `tp2`), you can discard this variable.
- `gesture`: type of gesture (`ho_gv` hold-out/give object, `point`ing with index finger, `reach` out with empty hand).
- `count_raw`: number of gestures without pro-rata correction, you can discard this variable.
- `count`: number of gestures.
- `ct_raw`: number of mother's contingent talks without pro-rata correction, you can discard this variable.
- `ct`: number of mother's contingent talks.
- `pro_rata`: whether a pro-rata adjustment was applied, you can discard this variable.
- `id`: ID of longitudinal trial.

This time we want to model the number of gesture `count`. The appropriate distribution family for counts in the Poisson family.

Models with a Poisson family return estimates in log-odds, similarly to what binomial/Bernoulli models do. While with binomial/Bernoulli estimates you use the `plogis()` function to go from log-odds to probabilities, you use the `exp()` function to convert log-odds to counts and odds.

Let's first create the necessary variables.

```{r gestures-ge-ba}
gestures <- gestures %>%
  mutate(
    dyad = as.factor(dyad),
    ge_ba = as.ordered(interaction(gesture, background))
  )
contrasts(gestures$ge_ba) <- "contr.treatment"
```

We can now fit a GAMM. Let's start simple.

```{r gest-gam}
gest_gam <- bam(
  count ~
    ge_ba +
    # We need to set k to 3 because k cannot be higher than the number of
    # unique values in the smoothing variable, here months.
    s(months, k = 3) + s(months, by = ge_ba, k = 3) +
    s(months, dyad, bs = "fs", m = 1, k = 3),
  data = gestures,
  family = poisson()
)
```

```{r gest-gam-sum}
summary(gest_gam)
```

Look at the parametric and smooth terms. What do they suggest?

Let's check the model with `gam.check()`.

```{r gest-gam-check}
gam.check(gest_gam)
```

Things look OK (the number of basis functions is good enough, although in any case we couldn't have done much since we cannot increase k).

Now time to plot! Fill the code below. Remember to specify the `tran_fun`, which needs to be `exp`, and to exclude the factor smooth. Feel free to pick comparison terms and facets.

```{r gest-gam-plot}
predict_gam(
  gest_gam,
  ...
) %>%
  plot(series = "months")
```

Great job!

Now a challenge: fit another GAMM as the one above, but now include `ct` as a smooth, and a tensor product interaction between `ct` and `months`.

```{r}

```

