---
title: "Hands-on 2"
author: "Stefano Coretta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
theme_set(theme_light())
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
library(mgcv)
options(show.signif.stars = FALSE)
library(tidygam)
```

## Compare across groups

### ACCH data

Let's expand on our analysis of the `acch` data.
In the first hands-on, we have looked at the effect of `Group` on logged reaction times.

We can now look at the effects of `Group`, `Structure` and `Number` and their combinations.

To do so, we first need to create a new column in `acch` with the factor interaction of `Group`, `Structure` and `Number`. Go ahead and fill in the following code to create a factor interaction. **Remember** that the factor has to be an ordered factor and that you also need to change the contrasts to `"contr.treatment"`.

```{r acch-gr-st-nu}
acch <- acch %>%
  mutate(
    Gr_St_Nu = ...
  )
...
```

Now you can fit a GAM with `Gr_St_Nu` as a `by`-variable to model the "interaction" between the levels of `Group`, `Structure` and `Number`.

Fill in the missing code below. Remember to include:

- A **parametric effect** of `Gr_St_Nu`.
- A **reference smooth** over Trial.
- A **difference smooth** over Trial **by** `Gr_St_Nu`.

```{r rt-gam-2}
rt_gam_2 <- gam(
  RT_log ~
    ...,
  data = acch
)
```

Now let's check the model summary.

```{r rt-gam-2-sum}
summary(rt_gam_2)
```

Look at the parametric coefficients. What do they indicate? Note that the Intercept is `bilingual.Object.Match` and that the estimate indicates differences in mean **height** of the smooths.

Look at the smooth terms, the Estimated Degrees of Freedom in particular. What do they indicate? Note that EDF > 1 means that the **shape** of that level or combination of levels is different from the reference smooth.

Let's inspect the model predictions.

```{r rt-gam-2-pred}
rt_gam_2_pred <- predict_gam(rt_gam_2)
```

To make plotting easier, let's separate the `Gr_St_Nu` factor interaction into the individual predictors.

```{r rt-gam-2-sep}
rt_gam_2_pred_sep <- predict_gam(
  rt_gam_2,
  # To separate factor interactions, use named lists of the form:
  # factor interaction = vector of variables to be separated into.
  separate = list(Gr_St_Nu = c("Group", "Structure", "Number")),
  length_out = 30
)
```

Let's plot now the predicted smooths. Feel free to try different comparisons/faceting.

```{r rt-gam-2-plot}
rt_gam_2_pred_sep %>%
  plot(series = "Trial", comparison = "Structure") +
  facet_grid(Number ~ Group)
```

### Electroglottographic (EGG) data from Italian and Polish speakers

For this exercise, you will use EGG data from Italian and Polish speakers, which contains time-series information on the closed quotient (info below).

```{r egg-cq}
egg_cq <- read_csv("data/egg_cq.csv")
```

The data consists of measurements of closed quotient taken from the entire duration of the first vowel in CVCV nonce words, as spoken by Italian and Polish speakers.

The closed quotient is a (relative) measure of glottal constriction (higher CQ values = greater glottal constriction).
CQ is also related to phonation: the lower the CQ the breathier the phonation.

The data contains time-series of CQ values in vowels (/a, o, u/) followed by voiceless or voiced consonants (/t, d, k, g/).
`rel_time` corresponds to the relative (proportional) time the measurement was taken from, within the duration of the vowel.

Go ahead and prepare the data so that you can fit a GAM over `rel_time`, while accounting for `language` (Italian vs Polish), `vowel` (/a, o, u/), `voicing` (voiced vs voiceless) and `place` (coronal vs velar).

First, create a factor interaction (make sure it's an ordered factor and that it has treatment contrasts).

```{r egg-cq-2}

```

Then fit the GAM, check the summary and produce plots based on the model.

```{r cq-gam}

```



## Tensor product smooths and interactions

