---
title: "Hands-on 3"
author: "Stefano Coretta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())
library(tidyverse)
theme_set(theme_light())
library(ggforce)
options(ggplot2.discrete.fill = RColorBrewer::brewer.pal(8, "Dark2"))
options(ggplot2.discrete.colour = RColorBrewer::brewer.pal(8, "Dark2"))
library(mgcv)
options(show.signif.stars = FALSE)
library(tidygam)
```

## Random effects

There are two main ways of including random effects in GAMs. Note that the concept of random effects is a tricky one, even in non-additive linear models and given the complexity of random effects as a concept in GAMs I will give you a general introduction on how to include random effects, but I refer you to Márton Sóskuthy's paper for a more in-depth treatment (<https://www.sciencedirect.com/science/article/pii/S009544702030108X>, the paper is on phonetics but applies to any field.)

Let's start with the data on f0 trajectories.

```{r vowels}
vowels <- read_csv("data/vowels.csv")
vowels
```

Look at the model without random effects and its plot below.

```{r f0-gam}
f0_gam <- gam(
  f0 ~ 
    ti(time_point) + ti(duration) + ti(time_point, duration),
  data = vowels
)
```

```{r f0-gam-plot-2, echo=TRUE}
predict_gam(f0_gam, values = list(duration = c(50, 100, 150, 200))) %>%
  plot(series = "time_point", comparison = "duration")
```

Now, let's add random effects.

### Random intercepts and random slopes

One way to include random effects is practically the GAM equivalent of random intercepts and random slopes in linear models.

Random intercepts allow the GAM to model different smooth heights for different participants, items, etc.
Random slopes allow the GAM to model different smooth rotations for different participants, items, etc.
Note that random intercept and slopes **do not** affect the **shape** of the smooth, just the height or its degree of rotation.

You can include random intercepts and slopes like so:

```{r f0-gam-2}
vowels <- vowels %>%
  mutate(
    # Variables you want to include as random effects must be factors (no need for them to be ordered).
    speaker = as.factor(speaker)
  )

f0_gam_2 <- bam(
  f0 ~ 
    ti(time_point) + ti(duration) + ti(time_point, duration) +
    # Random intercept by speaker
    s(speaker, bs = "re") +
    # Random slope of time point by speaker
    s(speaker, time_point, bs = "re"),
  data = vowels
)
```

```{r f0-gam-2-sum}
summary(f0_gam_2)
```

```{r f0-gam-2-plot}
predict_gam(
  f0_gam_2, values = list(duration = c(50, 100, 150, 200)),
  exclude_terms = c("s(speaker)", "s(speaker,time_point)")
  ) %>%
  plot(series = "time_point", comparison = "duration")
```

### Random effects with factor smooth interactions

An alternative way to include random effects is to use so-called **factor smooth interactions**.

Factor smooths interactions deal with height and rotation as the random intercept and slope do, but they also deal with the **shape** of the smooth, so you just need factor smooth interactions.

This is how ypu add factor smooth interactions.

```{r f0-gam-3}
# Remember that the variable you include in factor smooth interactions has to be a factor.

f0_gam_3 <- bam(
  f0 ~ 
    ti(time_point) + ti(duration) + ti(time_point, duration) +
    s(time_point, speaker, bs = "fs", m = 1),
  data = vowels
)
```

Notice that the order of the variable when including factor smooth interactions is the opposite of what you would do with random slopes. The basis function has to be `"fs"` for factor smooths and you also need to set `m` to 1 (this constraints the smoothing penalty further so that you don't over-fit the data).

Note that factor smooth interactions increase computation time but they are worth the wait.

```{r f0-gam-3-sum}
summary(f0_gam_3)
```

```{r f0-gam-3-plot}
predict_gam(
  f0_gam_3, values = list(duration = c(50, 100, 150, 200)),
  exclude_terms = c("s(time_point,speaker)")
  ) %>%
  plot(series = "time_point", comparison = "duration")
```

If you want to see the predicted f0 trajectory for some of the subjects, you can specify this with the `values` argument.

```{r f0-gam-3-plot}
predict_gam(
  f0_gam_3, values = list(
      duration = c(50, 100, 150, 200),
      speaker = c("it01", "it02", "it03", "pl03", "pl04", "pl05")
    )
  ) %>%
  plot(series = "time_point", comparison = "duration") +
  facet_wrap(~ speaker)
```
Let's see what happens when we also include a factor smooth interaction by-speaker for duration.

```{r f0-gam-4}
f0_gam_4 <- bam(
  f0 ~ 
    ti(time_point) + ti(duration) + ti(time_point, duration) +
    s(time_point, speaker, bs = "fs", m = 1) +
    s(duration, speaker, bs = "fs", m = 1),
  data = vowels,
  # The fast-REML method speeds model fitting
  method = "fREML"
)
```

You will notice that it took bit more to fit the model. But let's see what we got now.

```{r f0-gam-4-sum}
summary(f0_gam_4)
```



```{r f0-gam-4-plot}
predict_gam(
  f0_gam_4, values = list(duration = c(50, 100, 150, 200)),
  exclude_terms = c("s(time_point,speaker)", "s(duration,speaker)")
  ) %>%
  plot(series = "time_point", comparison = "duration")
```

Including factor smooths for time point and duration has increased the uncertainty around the estimated f0 trajectories. But the model (summary) still indicates that:

- The trajectory of f0 is not a straight line (`ti(time_point)` EDFs, p-value).
- Duration affects f0 in a non-linear way (`ti(duration)` EDFs, p-value).
- The "interaction" between time point and duration is non-linear (`ti(time_point, duration)` EDFs, p-value).
- Speaker specific variation is also non-linear (EDFs and p-values of factor smooths).

Let's take a peek at the predicted effects for six speakers.

```{r f0-gam-4-plot-2}
predict_gam(
  f0_gam_4, values = list(
      duration = c(50, 100, 150, 200),
      speaker = c("it01", "it02", "it03", "pl03", "pl04", "pl05")
    )
  ) %>%
  plot(series = "time_point", comparison = "duration") +
  facet_wrap(~ speaker)
```

Let's focus on the effect of vowel duration on f0 at vowel onset, mid-point and offset.

```{r f0-gam-4-plot-3}
predict_gam(
  f0_gam_4, values = list(
      time_point = c(1, 5, 9),
      speaker = c("it01", "it02", "it03", "pl03", "pl04", "pl05")
    ),
  length_out = 30
  ) %>%
  plot(series = "duration", comparison = "time_point") +
  facet_wrap(~ speaker)
```

Let's go one step further and let's look at the tensor product interaction in full.

```{r f0-gam-4-plot-4}
predict_gam(
  f0_gam_4, values = list(
      speaker = c("it01", "it02", "it03", "pl03", "pl04", "pl05")
    )
  ) %>%
  plot(series = c("time_point", "duration"), raster_interp = TRUE) +
  facet_wrap(~ speaker)
```

You see that some types of plots are easier to interpret than others, so it is always up to you to decide which one to include in a paper depending on what message you want to convey.
